---
title: "BCB420 Assignment 1: GEOMetaDB Searching"
author: "Veronica Chang"
date: "February 14th, 2023"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: sentence
bibliography: a1.bib
---

# Install packages and select an expression data set

## Packages
* [BiocManager](https://CRAN.R-project.org/package=BiocManager) [@morgan2022]
* [GEOmetadb](https://bioconductor.org/packages/release/bioc/html/GEOmetadb.html) [@zhu2008]
* [limma](https://bioconductor.org/packages/release/bioc/html/limma.html) [@smyth2015]
* [edgeR](https://bioconductor.org/packages/release/bioc/html/edgeR.html) [@robinson2010]
* [biomaRt](https://bioconductor.org/packages/release/bioc/html/biomaRt.html) [@durinck2009]
* [GEOquery](https://bioconductor.org/packages/release/bioc/html/GEOquery.html) [@davis2007]
* [Biobase](https://bioconductor.org/packages/release/bioc/html/Biobase.html) [@huber2015]
* [RColorBrewer](https://cran.r-project.org/package=RColorBrewer) [@neuwirth2022]
```{r, message=FALSE, warning=FALSE}

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("GEOmetadb", quietly = TRUE))
    BiocManager::install("GEOmetadb")

library("BiocManager")
library("GEOmetadb")
library("limma")
library("edgeR")
library("dplyr")
library("Biobase")
library("biomaRt")
library("GEOquery")
library("RColorBrewer")
library("ggplot2")
library("knitr")
library("RSQLite")
library("kableExtra")
```


## Prepare the database

```{r, message=FALSE, tidy=TRUE}
# Check if it already exists
if(!file.exists('GEOmetadb.sqlite')) GEOmetadb::getSQLiteFile()
file.info('GEOmetadb.sqlite')

# Connect to the database
con <- DBI::dbConnect(RSQLite::SQLite(),'GEOmetadb.sqlite',`synchronous` = NULL)
DBI::dbListFields(con, 'gpl')
geoTables <- DBI::dbListTables(con)
geoTables
```


## Query the database

I am looking for series related to the human autoimmune diseases, lupus or multiple sclerosis. Therefore, the organisms should contain `Homo sapiens` and the title of the series should contain one of `multiple sclerosis` or  `lupus`. Technology should be `Highthroughoutput RNA sequencing` and was submitted within the past `10 years`.

```{r, message=FALSE, tidy=TRUE}
results <- DBI::dbGetQuery(con,'select * from gpl limit 5')
knitr::kable(head(results[,1:5]), format = "pipe")

# Build the SQL query
sql <- paste("SELECT DISTINCT gse.title, gse.gse, gpl.title,",
             " gse.submission_date,",
             " gse.supplementary_file",
             "FROM",
             "  gse JOIN gse_gpl ON gse_gpl.gse=gse.gse",
             "  JOIN gpl ON gse_gpl.gpl=gpl.gpl",
             "WHERE",
             "  gpl.organism LIKE '%Homo sapiens%' AND",
             "  gse.title LIKE '%lupus%' OR gse.title LIKE '%multiple sclerosis%' AND",
             "  gpl.technology LIKE '%high-throughput sequencing%' AND",
             "  gse.submission_date > '2013-01-01'",
             "  ORDER BY gse.submission_date DESC",
             sep=" ")

# Initiate the query
rs <- DBI::dbGetQuery(con,sql)
knitr::kable(rs, format = "pipe")
dim(rs)
```