---
title: "BCB420 Assignment 1: Data set selection and initial processing."
author: "Veronica Chang"
date: "`r format(Sys.time(), '%d %b %Y')`"
output:
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: sentence
---

# Task 1: Install packages and select an expression data set.

```{r, message=FALSE, warning=FALSE}

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("GEOmetadb", quietly = TRUE))
    BiocManager::install("GEOmetadb")
```

Prepare the database

```{r, message=FALSE}
# Check if it already exists
if(!file.exists('GEOmetadb.sqlite')) GEOmetadb::getSQLiteFile()
file.info('GEOmetadb.sqlite')

# Connect to the database
con <- DBI::dbConnect(RSQLite::SQLite(),'GEOmetadb.sqlite',`synchronous` = NULL)
DBI::dbListFields(con, 'gpl')
geoTables <- DBI::dbListTables(con)
geoTables
```

Going to query the database.
I am looking for series related to the human autoimmune disease, lupus or SLE.
Therefore, the organims should contain `Homo sapiens` and the title of the series should contain one of `autoimmune`, `lupus`.
Technology should be `Highthroughoutput RNA sequencing` and was submitted within the past `10 years`.

```{r}
results <- DBI::dbGetQuery(con,'select * from gpl limit 5')
knitr::kable(head(results[,1:5]), format = "pipe")

# Build the SQL query
sql <- paste("SELECT DISTINCT gse.title, gse.gse, gpl.title,",
             " gse.submission_date,",
             " gse.supplementary_file",
             "FROM",
             "  gse JOIN gse_gpl ON gse_gpl.gse=gse.gse",
             "  JOIN gpl ON gse_gpl.gpl=gpl.gpl",
             "WHERE",
             "  gpl.organism LIKE '%Homo sapiens%' AND",
             "  gse.title LIKE '%lupus%' OR gse.title LIKE '%multiple sclerosis%' AND",
             "  gpl.technology LIKE '%high-throughput sequencing%' AND",
             "  gse.submission_date > '2013-01-01'",
             "  ORDER BY gse.submission_date DESC",
             sep=" ")

# Initiate the query
rs <- DBI::dbGetQuery(con,sql)
knitr::kable(rs, format = "pipe")
dim(rs)
```

# Task 2: Clean the data and map to HUGO symbols.

```{r}

```

# Task 3: Apply Normalization.

```{r}

```

# Task 4: Interpret, and document.

What are the control and test conditions of the dataset?
Why is the dataset of interest to you?
Were there expression values that were not unique for specific genes?
How did you handle these?
Were there expression values that could not be mapped to current HUGO symbols?
How many outliers were removed?
How did you handle replicates?
What is the final coverage of your dataset?
